ranacapa: An R package to explore environmental DNA data with exploratory statistics and interactive visualizations

### 2018-10-05 manuscript.Rmd
adds comments and some fixes

The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA (eDNA) metabarcoding",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].  Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, ranging from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017]. As the cost of eDNA metabarcoding declines and sample collection techniques become more streamlined (e.g. @Thomas_2018), professional research scientists are increasingly using eDNA metabarcoding as a platform to engage a diversity of community partners, including natural resource managers, undergraduate students, and citizen scientists in primary research. However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.

@@

To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp that can be used to visualize results from eDNA sequencing studies and perform simple community ecology analyses. `ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations of several core analyses used in eDNA studies core and includes links to additional educational resources. `ranacapa` works with community matrices generated via QIIME [Caporaso_2010] or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), which is used extensively by the CALeDNA program.

In the remainder of this manuscript, we describe the package and how it is used by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA. As we show in the Use Cases, empowering community partners to interact with the data and perform simple but insightful community ecology analyses can help make these collaborations more enriching and valuable to both parties.

## Implementation
At the core of `ranacapa` is a Shiny webapp (@Chang_2018), which is avialable at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`. The package also includes two categories of helper functions (Table 1) that transform user-uploaded taxonomy and metadata tables into `R` objects that can be visualized and analyzed using the Phyloseq [@McMurdie_2013] and Vegan [@Oksanen_2018] packages. `ranacapa` is available for installation from Github or CRAN:

@@

- **Beta diversity plots**: Introduces the concept of Beta diversity as the turnover in species composition across habitats (or samples). The tab includes an ordination plot generated by `phyloseq::plot_ordination()`, which in turn uses as ordination made with `phyloseq::ordinate(., method = "PCoA")`. Points on the PCoA plot are colored according to a user-selected metadata variable (Figure 5).

@@

- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site. The first statistical test is a multivariate ANOVA implemented with `vegan::adonis())`, which shows results from a statistical test of taxon turnover across sites. The second statistical test is of heterogeneity of variances among samples implemented with `vegan::betadisper()`, dthat compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).

@@

A team of UCLA researchers partnered with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement expensive and time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996; @Usseglio_2015; @Deiner_2017]. For this partnership, resource managers collected and filtered thirty-1L water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S and CO1 gene, targeting bony fishes, elasmobranches, and invertebrate taxa. The researchers processed sequences and assigned taxonomy using the Anacapa pipeline, and shared results with the resource managers using the `ranacapa` Shiny app.

The taxonomy heatmap (Figure 2) was the most valuable visualization to this collaboration, because it allowed the resource managers to filter the large observed species list down to a particular set of key taxa that they regularly monitor. The heatmap showed that this pilot study detected 36 of the 70 key metazoans at the species level, and the remaining 34 at the genus, family, or order level. This indicates that eDNA-based studies can likely supplement ongoing management efforts and provide new insights into the spatial and temporal distributions of these species. The value of `ranacapa` in this scenario was to quickly sort through long species lists generated in by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands. The data from this study are packaged as the demo dataset for the `ranacapa` Shiny web-app and are available online at XXX.

@@

A team of community ecology and environmental DNA researchers collaborate with instructors of a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community. The goal of this twenty-week course was to provide undergraduate students an authentic experience in basic microbiology and microbial community ecology research. Over the first ten weeks, eDNA researchers on the instructional team sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding regions from student=collected soil samples and used the Anacapa pipeline to generate taxon-by-sample tables.

>>>>

The targeted amplification and sequencing of DNA that living organisms shed into their physical environment, termed "environmental DNA (eDNA) metabarcoding",  is revolutionizing microbiology, ecology, and conservation research [@Taberlet_2012; @Deiner_2017].  Sequencing of environmental DNA extracted from field-collected soil, water, or sediment samples can yield insight into a range of questions, from profiling the composition of ancient plant and animal communities [@Pedersen_2014], to motoring populations of rare or endangered species [@Balasingham_2017]. As the cost of eDNA metabarcoding declines and sample collection techniques become more streamlined (e.g. @Thomas_2018), professional research scientists are increasingly using eDNA metabarcoding as a platform to engage a diversity of community partners, including natural resource managers, undergraduate students, and citizen scientists in primary research. However, developing robust and impactful community science programs that engage community partners in all steps of the research process remains a challenge.

@@

To address this challenge, we created an R package “ranacapa”, at the core of which is a Shiny webapp that can be used to visualize results from eDNA sequencing studies and perform simple community ecology analyses. `ranacapa` complements existing visualization platforms (e.g. Phinch, Phyloseq-Shiny, QIIME2 Viewer), because in addition to interactive visualizations, `ranacapa` includes brief explanations of several core analyses used in eDNA studies and includes links to additional educational resources. `ranacapa` works with community matrices generated via QIIME [Caporaso_2010] or the Anacapa sequence analysis pipeline (https://github.com/limey-bean/Anacapa), the latter being used extensively by the CALeDNA program.

Here, we describe the package and how it is used by two community science partnerships based at the University of California, Los Angeles (UCLA): first, a collaboration between eDNA researchers and resource managers at the National Park Service, and second, a partnership between community ecology researchers and an undergraduate microbiology course at UCLA. As we show in the Use Cases, empowering community partners to interact with the data and perform simple but insightful community ecology analyses can help make these collaborations more enriching and valuable to both parties.

## Implementation
At the core of `ranacapa` is a Shiny webapp (@Chang_2018), which is available at http://gauravsk.shinyapps.io/ranacapa or with `ranacapa::runRanacapa()`. The package also includes two categories of helper functions (Table 1) that transform user-uploaded taxonomy and metadata tables into `R` objects that can be visualized and analyzed using the Phyloseq [@McMurdie_2013] and Vegan [@Oksanen_2018] packages. `ranacapa` is available for installation from Github or CRAN:

@@

- **Beta diversity plots**: Introduces the concept of beta diversity as the turnover in species composition across habitats (or samples). The tab includes an ordination plot generated by `phyloseq::plot_ordination()`, which in turn uses an ordination object made with `phyloseq::ordinate(., method = "PCoA")`. Points on the PCoA plot are colored according to a user-selected metadata variable (Figure 5).

@@

- **Beta diversity statistics**: Shows results from two statistical tests of species turnover across site. The first statistical test is a multivariate ANOVA implemented with `vegan::adonis())`, which shows results from a statistical test of taxon turnover across sites. The second statistical test, which is implemented with `vegan::betadisper()`, is of heterogeneity of variances among samples. This test compares the degree of sample-to-sample variation within habitats (or within other user-selected groups).

@@

A team of UCLA researchers partnered with resource managers at the Channel Islands National Park Service to assess the potential for eDNA as a biodiversity monitoring tool to supplement expensive and time-intensive visual biodiversity surveys in the Southern California Channel Islands [@Lessios_1996; @Usseglio_2015; @Deiner_2017]. For this partnership, resource managers collected and filtered thirty unique one-liter water samples for eDNA analysis at permanent monitoring sites inside and adjacent to protected areas, and research scientists at UCLA performed eDNA sequencing of the mitochondrial 12S [@Miya_2015] and CO1 [Leray_2013] genes, targeting bony fishes, elasmobranches, and invertebrate taxa. The researchers processed sequences and assigned taxonomy using the Anacapa pipeline, and shared results with the resource managers using the `ranacapa` Shiny app.

The taxonomy heatmap of species detected using the 12S and CO1 metabarcodes (Figure 2) was the most valuable visualization to this collaboration, because it allowed the resource managers to filter the large observed species list down to a particular set of key taxa that they regularly monitor. The heatmap showed that this pilot study detected 36 of the 70 key metazoans at the species level, and the remaining 34 at the genus, family, or order level. This indicates that eDNA-based studies can provide critical information for ongoing management efforts and provide new insights into the spatial and temporal distributions of these species. The value of `ranacapa` in this scenario was to quickly sort through long species lists generated in by eDNA sequencing to highlight the strengths and weaknesses in using eDNA to monitor diversity in the Channel Islands. The data from this study are packaged as the demo dataset for the `ranacapa` Shiny web-app and are available online at XXX.

@@

A team of community ecology and environmental DNA researchers in the CALeDNA program collaborate with instructors of a research-based environmental microbiology course at UCLA [@Shapiro_2015], in which students used eDNA metabarcoding to study the impact of a local wildfire on the plant and soil microbial community. The goal of this twenty-week course was to provide undergraduate students an authentic experience in basic microbiology and microbial community ecology research. Over the first ten weeks, eDNA researchers on the instructional team sequenced the ITS2 [@Gu_2013] and 16S SSU RNA [@Caporaso_2012] metabarcoding regions from student=collected soil samples and used the Anacapa pipeline to generate taxon-by-sample tables.
